---
title: "Female Singles in DK"
author: "Nicole Dwenger"
date: "3/25/2021"
output: html_document
#editor_options: 
#  chunk_output_type: console
---

## Task: Autocorrelation of Distribution of single women in Denmark during 2020

Test for the autocorrelation of the distribution of single women OR single men in Denmark during 2020, and answer the question: "Is the population of single women/men in Denmark spatially correlated? What is the correlation and how significant is the trend?. You can download the attribute data either from Denmark Statistic, or find a slightly tidied dataset at this [URL](https://docs.google.com/spreadsheets/d/1xcrd07gV3Sm0fuzSIWu2Op36oDBmvvrlHU9uNz49kuU/edit#gid=0).

### Getting libraries
```{r libraries, message = FALSE}
# libraries
library(raster)
library(rgeos)
library(sf)
library(tidyverse)
library(htmltools)
library(googlesheets4)
library(cartogram)
library(mapview)
library(tmap)
library(spdep)
library(here)
library(knitr)
library(rmdformats)
```

### Getting and cleaning spatial data
```{r spatial data , message = FALSE}
# loading data and turning into sf
mun_sp <- readRDS(here("Week08", "data", "gadm36_DNK_2_sp.rds"))
mun_sf <- st_as_sf(mun_sp)
mun <- st_transform(mun_sf, crs = 32632)
# quick view
#mapview::mapview(mun)

# fixing names
mun$NAME_2[31] <- "Aarhus"
mun$NAME_2[21] <- "Høje-Taastrup"
mun$NAME_2[60] <- "Vesthimmerlands"
```

### Getting and cleaning relationship data
```{r singles data, message = FALSE}
# loading the data
#relationships <- read_sheet("https://docs.google.com/spreadsheets/d/1xcrd07gV3Sm0fuzSIWu2Op36oDBmvvrlHU9uNz49kuU/edit#gid=0")
#write_csv(relationships, "Week08/HW08/relationships.csv")
relationships <- read_csv(here("Week08", "HW08", "relationships.csv")) %>% na.omit()

# 104 regions
colnames(relationships)
relationships %>% group_by(Sex,Status) %>%
  summarise(n = n())

# 99 regions in mun, 104 regions in relationship data, so I'm removing them
out_regions <- setdiff(relationships$Region, mun$NAME_2)
relationships <- relationships %>% filter(!Region %in% out_regions)
length(unique(relationships$Region)) # now 99
```

### Preprocessing female data of 2020
```{r female percentages, message = FALSE}
# filtering on females in 2020
f_2020 <- relationships %>% 
  # filter on only women
  filter(Sex == "Kvinder") %>% 
  # create new column with TRUE/FALSE for single/not single
  mutate(Single = ifelse(Status == "Gift/separeret", FALSE, TRUE)) %>%
  # group by region and single column
  group_by(Region, Single) %>%
  # calculate sums, i.e. how many are single, how many are not single 
  summarize(N2020 = sum(Y2020K1))

# calculate "population", i.e. the sum of all women for each region
f_totals <- f_2020 %>%
  # grouping by region
  group_by(Region) %>%
  # calculating the sum of numbers, i.e. total population
  summarize(sum2020 = sum(N2020))

# calculate the percentages of someone being single or not 
f_percentages <- f_2020 %>%
  # adding the totals 
  merge(f_totals, by = "Region") %>%
  # adding column for percentages of singles/not singles
  mutate(pct_2020 = N2020/sum2020*100)
  
# merging the spatial data
f_data <- mun %>%
  merge(f_percentages, by.x = "NAME_2", by.y="Region")

```

### Mapping
```{r mapping}

# getting only singles data
f_singles <- f_data %>% filter(Single==TRUE)

# quick map of % of female singles in 2020 in Denmark
f_singles %>%
  group_by(NAME_2) %>%
  select(pct_2020) %>%
  mapview(layer.name = "% of Single Females")
```

### Catogram
```{r, message = FALSE}
# scatter plot of area and total population 
plot(f_singles$sum2020, st_area(f_singles, byid = TRUE))
# scatter plot of area and percentage of singles
plot(f_singles$pct_2020, st_area(f_singles, byid = TRUE))

# cartogram, scaling area to total population
cartogram_sum2020 <- cartogram_cont(f_singles, "sum2020")

# check linearity
plot(cartogram_sum2020$sum2020, st_area(cartogram_sum2020, byid = TRUE))

# cartogram, scaling area to percentage of singles
cartogram_singles2020 <- cartogram_cont(f_singles, "pct_2020")

# Check the linearity of the DF voter percentage per municipality plot
plot(cartogram_singles2020$pct_2020, st_area(cartogram_singles2020, byid = TRUE))

# fairer map of female single percentage in 2020
plot(cartogram_singles2020$geometry, 
     col = "beige",
     main = "% of Single Females in DK")
```

### Autocorrelation Test
#### Based on queen adjacency
```{r}
# simplifying geometry of municipalities
mun_sm <- st_cast(st_simplify(mun, dTolerance = 250), to = "MULTIPOLYGON")
plot(mun_sm$geometry)
# getting neighbors based on queen adjacency
nb_queen <- poly2nb(mun_sm$geometry)
nb_queen
# getting centers of the municipalities
mun_centers <-st_centroid(mun_sm$geometry, of_largest_polygon = TRUE)
# plotting connections
plot(mun_sm$geometry); plot(nb_queen, mun_centers, col = "red",add = TRUE)

# morans test: not significant
moran.test(f_singles$pct_2020, nb2listw(nb_queen, style = "W",zero.policy=TRUE),zero.policy=TRUE)
# monte carlo simulation: not significant 
MC_queen = moran.mc(f_singles$pct_2020,nb2listw(nb_queen, zero.policy=TRUE),zero.policy=TRUE, nsim = 999)
MC_queen
# plot MC simulations
plot(MC_queen, main="Distribution of MC simulation based on queen neighbours", las=1)
```

#### Based on 100km distance 
```{r}
# make neighbor list from neighbours at 100km distance
nb_100 <- dnearneigh(mun_centers, 0, 100000)
plot(mun_sm$geometry); plot(nb_100, mun_centers, col = "red",add = TRUE);title(main="Neighbours with 100km distance")

# morans test (using "less" because the value is negative): not significant 
moran.test(f_singles$pct_2020, nb2listw(nb_100, style = "W",zero.policy=TRUE),zero.policy=TRUE, alternative="less")
# monte carlo simulation: not significant 
MC_100 = moran.mc(f_singles$pct_2020,nb2listw(nb_100, zero.policy=TRUE),zero.policy=TRUE, nsim = 999, alternative="less")
MC_100
# plot MC simulations
plot(MC_100, main="Distribution of MC simulation based on 100km distance neighbours", las=1)
```

#### Based on 50km distance
```{r}
# make neighbor list from neighbours at 50km distance
nb_50 <- dnearneigh(mun_centers, 0, 50000)
plot(mun_sm$geometry); plot(nb_50, mun_centers, col = "blue",add = TRUE);title(main="Neighbours within 50km distance")

# morans test: not significant
moran.test(f_singles$pct_2020, nb2listw(nb_50, style = "W",zero.policy=TRUE),zero.policy=TRUE, alternative ="less")
# monte carlo simulation: not significant 
MC_50 = moran.mc(f_singles$pct_2020,nb2listw(nb_50, zero.policy=TRUE),zero.policy=TRUE, nsim = 999, alternative="less")
MC_50
# plot MC simulations
plot(MC_50, main="Distribution of MC simulation based on 50km distance neighbours", las=1)
```

#### Based on k-neighbours
```{r}
# getting the three neirest neighbours
mat <- as.matrix(st_coordinates(mun_centers))
k3 <- knearneigh(mat, k=3, longlat=FALSE)
nb_k3 <- knn2nb(k3)
plot(nb_k3, mun_sm$geometry)

# morans test: not significant
moran.test(f_singles$pct_2020, nb2listw(knn2nb(k3), style = "W",zero.policy=TRUE),zero.policy=TRUE)
# monte carlo simulation: not significant
MC_k3 <- moran.mc(f_singles$pct_2020,nb2listw(knn2nb(k3), zero.policy=TRUE),zero.policy=TRUE, nsim = 999)
MC_k3
# plot MC simulations
plot(MC_k3, main="Distribution of MC simulation based on k=3 neighbours", las=1)
```

### Conclusion
Testing spatial correlation with different based on different neighborhoods did not reveal any spatial clustering. For some tests Moran's I was negative, while it was positive for others. However, overall none of the values were significant (all > .05), suggesting that the spatial clustering was not significantly different from a random clustering. 